Light and growth form interact to shape stomatal ratio among British angiosperms

### 2017-07-14 ms/ms.Rnw
edited ms

% Things to do before submission:
% Remove code to copy refs
% Remove notes
% word count

@@

\usepackage{geometry}
\geometry{letterpaper}
\usepackage[parfill]{parskip}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{topcapt}
\usepackage{caption}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{natbib}
\usepackage{color}
\usepackage{array}
\usepackage{gensymb}
\usepackage{setspace}
\newcommand{\stretchy}{2}
\usepackage{lineno}
\usepackage{array}
\usepackage{lscape}
\usepackage{multirow}

@@

\title{Light and life form interact to shape stomatal ratio among British angiosperms}
\author{Christopher D. Muir$^1$}
\date{} % delete this line to display the current date

@@

$^1$ Biodiversity Research Centre and Botany Department, University of British Columbia, Vancouver, British Columbia V6T 1Z4, Canada \\
\\
\textit{Author for correspondence:} \\
\textit{Christopher D. Muir} \\
\textit{Tel:} +17782284851 \\
\textit{Email:} chrisdmuir@gmail.com \\
University of British Columbia \\
6270 University Blvd. \\
Vancouver, BC, Canada \\
V6T 1Z4 \\
\\
Short title: Shedding light on stomatal evolution\\
\\
Word count: \\
Summary:  193\\
Introduction:  \\
Materials and Methods:  \\
Results:  \\
Discussion:  \\
Acknowledgement:  \\
3 Figures and 1 Table, 1 Supplemental Figure

@@

  # Remove before publication
  #file.copy(from = "~/Google Drive/CommonResources/refs.bib",
  #          to = "~/Google Drive/britstom/ms/refs.bib", overwrite = TRUE)

@@

\begin{itemize}
	\item In most plants, stomata are located only on the abaxial leaf surface (hypostomy), but many plants have stomata on both surfaces (amphistomy). High light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together using phylogenetic comparative methods.
	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, Raunki\ae r lifeform, and phylogenetic relationships for \Sexpr{nrow(stomata)} species of British angiosperms. I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
	\item High light and herbaceous growth form are correlated with amphistomy, as predicted, but they also interact; the effects of light are most pronounced in therophytes (annuals) and perennial herbs, but less so in phanerophytes (mostly trees). Interestingly, amphistomy and stomatal density evolve together in response to light, suggesting coordinated selection on this trait combination.
	\item I show for the first time that light and growth form interact to shape variation in stomatal ratio; amphistomy is advantageous in high light, but mostly for herbs. These results improve our understanding of the adaptive significance of stomatal ratio as well as its use as functional trait for paleovegetation reconstruction and crop improvement.
\end{itemize}

@@

Adaptation, amphistomy, Ellenberg light indicator value, growth form, phylogenetic comparative methods, stomata, stomatal ratio

@@

The leading, nonmutually exclusive hypotheses are that 1) open habitats favour amphistomy because CO$_2$ diffusion most strongly limits photosynthetic rate under high light and 2) herbaceous growth form favours amphistomy because traits that maximize photosynthetic rate are often under stronger selection in herbs. \citeauthor{Salisbury_1927} (\citeyear{Salisbury_1927}) first noted that amphistomy is most common in herbaceous plants from open habitats (i.e., with high light) of the British flora. These observations have been replicated in other studies \citep{Mott_etal_1982, Peat_Fitter_1994b, Jordan_etal_2014, Muir_2015} and may support physiological and ecological hypotheses that CO$_2$ most strongly limits photosynthesis in high light and/or photosynthesis contributes most to fitness in herbaceous plants. Under high light, CO$_2$ can strongly limit maximum photosynthetic rates, espcecially in thick leaves \citep{Jones_1985}. Hence, having stomata on both surfaces relieves this limitation by adding a second parallel pathway for CO$_2$ diffusion. Parkhurst \citeyear{Parkhurst_1978} argued that greater leaf thickness \textit{per se} selected for amphistomy, but there is little evidence for correlations between leaf thickness and stomatal ratio independent of light \citep{Mott_etal_1982, Gibson_1996, Muir_2015}. Amphistomy is correlated with open habitat in warm desert plants of western North America \citep{Mott_etal_1982, Gibson_1996}, among the Proteaceae \citep{Jordan_etal_2014}, and in continental European herbs \citep{Bucher_etal_2017}.

Stomatal ratio is also associated with growth form. In the British flora, \citeauthor{Salisbury_1927} (\citeyear{Salisbury_1927}) found that trees and shurbs are nearly always hypostomatous, whereas herbs from open habitats are amphistomatous. This pattern holds when data are averaged by family to coarsely control for phylogenetic nonindependence \citep{Peat_Fitter_1994b} or when using alternative classification schemes, such as Raunki\ae r life form \citep{Peat_Fitter_1994b}. Across plants from $\sim 90$ familes worldwide, growth form is the strongest predictor of stomatal ratio when multiple factors are estimated simultaneously and controlling for phylogenetic nonindependence \citep{Muir_2015}. These patterns are consistent with other data indicating that many herbaceous plants are under strong selection for high maximum photosynthetic rates. \citep{Bazzaz_1979, Korner_etal_1989, Wullschleger_1993}.

@@

The final question is important for telling whether amphistomy is part of a coordinated syndrome of traits that promote higher photosynthetic rate, as both the light and growth form hypotheses assume. If evolved increases in stomatal ratio are mediated by shifting abaxial stomata to the adaxial surface, holding total stomatal density constant, then the overall increase in CO$_2$ diffusion would be limited. In contrast, if amphistomy evolves by increasing adaxial stomatal density while holding abaxial density constant, then \textit{total} stomatal density must increase as well. Evolutionary coordination of amphistomy and high stomatal density would reinforce one another, increasing CO$_2$ supply to chloroplasts more than changes in either trait would in isolation.

@@

I obtained data on ab- and adaxial stomatal density on \Sexpr{nrow(stomata_unfiltered)} species from British Ecological Flora \citep{Salisbury_1927, Fitter_Peat_1994a, BEF}. Following recent comparative analyses \citep[e.g.][]{Bartelheimer_Poschlod_2016, Salguero-Gomez_etal_2016}, I used Ellenberg light indicator values \citep{Ellenberg_1974} and Raunki\ae r life form \citep{Raunkiaer_1934} as measures of light habitat and growth form, respectively. Hence, I am assuming that the species' light habitat is closely related to the type of habitat (open versus closed) where that species is found. Both attributes have been recently updated by taxonomic experts of the British flora (PLANTATT, \cite{Hill_etal_2004}). Ellenberg light indicator values are hereafter abbreviated \el. I used a dated molecular phylogeny of the British flora \citep{Lim_etal_2014} available from TreeBASE (http://treebase.org/; accession number 15105). \Sexpr{nNotInPhy1} species (\Sexpr{round(100 * nNotInPhy1 / nrow(stomata_unfiltered), 1)}\%) in the dataset were not present in the phylogeny. For \Sexpr{nNotInPhy1 - nNotInPhy2} of these species, I used the position a congeneric species as a proxy for the focal species. When multiple congeneric species were present, I consulted the phylogenetic literature to identify the most closely related proxy species \citep{Scheen_etal_2004, Salmaki_etal_2013}. For the remaining \Sexpr{nNotInPhy2} missing species, I positioned them in the tree based on phylogenetic relationships to other genera or families present in the tree \citep{Fior_etal_2006}. Because many phylogenetic comparative methods do not allow polytomies, zero-length branches, and non-ultrametric trees, I made several small adjustments to the tree. I resolved polytomies randomly using the `multi2di' function in \pkg{phytools} version 0.6-00 \citep{Revell_2012}. I added 0.02 my to all zero-length branches, as this was approximately the length of the shortest nonzero branch length in the tree. After these changes, I slightly altered terminal branch lengths to make the tree precisely ultrametric.

@@

I compared phylogenetic linear models to test whether Raunki\ae r life form, \el, or interactions between them predicted $\textrm{SR}_\textrm{even}$. I used $\textrm{SR}_\textrm{even}$ rather than $\textrm{SR}_\textrm{propAd}$ as the response variable because the hypothesis is that faster life history and/or high light favor more even stomatal densities on each surface. I fit models using \pkg{phylolm} and extracted Akaike Information Criteria (AIC). For these and subsequent analyses, I assumed an Ornstein-Uhlenbeck process model for the residuals with the root character state integrated over the stationary distribution. I used a 10$^4$ parametric bootstrap samples of the full model (including main effects and interactions) to calculate parameter confidence intervals \citep{Boettiger_etal_2012}. % REMOVING, but keep text in case I decide to replace: Likewise, to determine whether the interaction between Raunki\ae r life form and \el~was statistically significant, I used a parametric bootstrap to generate the null distribution of $\Delta$AIC values ($\Delta$AIC is the difference in AIC between competing models). Specifically, I sampled 1000 random datasets from the estimated model with main effects of Raunki\ae r life form and \el~but no interaction. I fit these simulated datasets to models with and without interactions and  calculated $\Delta$AIC. The statistical significance of the observed $\Delta$AIC is the proportion of simulated $\Delta$AIC greater than the observed.

@@

I used two related phylogenetic methods, variance decomposition and structural equation modeling (SEM), to assess the relative contribution of ab- versus adaxial stomatal density to light-mediated stomatal ratio evolution. First, the contribution of each ab- versus adaxial stomatal density can be calculated using phylogenetic variance decomposition methods as derived below. Because stomatal density is highly skewed, I log-transformed values for normality:

@@

I also wanted to test whether light-mediated evolution of stomatal ratio acted mostly by 1) increasing adaxial stomatal density while maintaining abaxial density, or 2) keeping total stomatal density the same, but shifting a greater proportion to the adaxial surface. The first scenario predicts that the phylogenetic regression of \el~ on $\mathrm{sd_{ad}}$ is stronger than that for $\mathrm{sd_{ab}}$. The second scenario predicts that \el~acts similarly on both and that there is a negative covariance $\mathrm{Cov(sd_{ad}, sd_{ab}) < 0}$. I tested these competing predictions by fitting a very simple phylogenetic SEM. The model uses the phylogenetic covariance matrix, as described above, to simultaneously estimate regressions of \el~on $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ while allowing covariance between them (i.e. estimating $\mathrm{Cov(sd_{ad}, sd_{ab})}$). To fit the SEM, I used the R package \pkg{lavaan} version 0.5-23.1097 \citep{Rosseel_2012}. I tested whether parameter estimates were significantly different from zero using $z$-scores.

@@

The ratio of stomatal densities on the abaxial (`lower') to that of the adaxial (`upper') surface varies greatly across plant species, but the adaptive significance is not clear. Comparative studies correlating stomatal ratio to ecological factors can distinguish among competing hypotheses and reveal critical experiments for future work. Previous comparative studies suggested that high light and herbaceous growth form favor amphistomy \citep{Mott_etal_1982, Jordan_etal_2014, Muir_2015, Bucher_etal_2017}, particularly in the British flora \citep{Salisbury_1927, Peat_Fitter_1994b}. However, none of these studies have accounted for the fact thats light and growth form are often confounded -- open, high light habitats are often dominated by herbs -- or the fact that species are not independent because of shared evolutionary hisory. Here, I reanalyzed data on stomata, light tolerance, and growth form in British angiosperms using phylogenetic comparative methods. As expected, species' light tolerance (Ellenberg light indicator or \el) is confounded with growth form (Raunki\ae r life form; Fig.~\ref{fig:lf-light}). Nevertheless, both \el~and Raunki\ae r life form affect stomatal ratio, but these factors also interact; the influence of \el~on stomatal ratio varies across forms. These novel findings provide further evidence that variation in stomatal ratio is adaptive and have important implications for interpreting changes in stomatal ratio through the paleo record \citep{Jordan_etal_2014} and during domestication \citep{Milla_etal_2013}.

@@

Previously, associations between light, growth form, and stomatal ratio have been interpreted in isolation as indicating that either high light and/or herbaceous growth form favors amphistomy. In British angiosperms, both factors are important, though statistical analyses suggest that light may be a stronger determinant than growth form (Table~\ref{table:srAIC}). Unlike previous studies, I found a significant interaction between light and growth form among British angiosperms, which suggests that amphistomy may only be strongly favored when CO$_2$ strongly limits photosynthesies (as in open habitat) \textit{and} photosynthesis strongly limits fitness (as in herbs). This is consistent with life history theory predicting that the demography of open habitat herbs is strongly limited by plant growth \citep{Franco_Silvertown_1996}. The ideal way to test this would be to measure selection on stomatal ratio in a species that varied quantitatively in both stomatal ratio and life history (e.g., containing both annual and perennial forms). I predict that amphistomy will be favored more strongly in the annual form grown under high light compared to an annual under low light or a perennial in high light, and much more strongly than a perennial grown in low light. Similar experiments could also be performed to test if and when light-mediated plasticity in stomatal ratio is adaptive \citep{Gay_Hurd_1975, Mott_Michaelson_1991}.

@@

The nonsignificant relationship between \el~and stomatal ratio in geophytes and phanerophytes suggests that in some cases amphistomy may not reliably indicate open habitat without further information. For example, perhaps amphistomatous geophytes from partially shaded habitats are spring ephemerals, so they  experience high light during their growth phase, but this has not been tested. Likewise, phanerophytes (most tall trees) are almost always hypostomatous (see also \cite{Muir_2015}). Most British phanerophytes are tall, hypostomatous trees, but the exceptions are telling. For example, the most amphistomatous phanerophyte in this dataset is \textit{Brassica oleracea}, a short-statured biennial that has more in common physiologically with hemicryptophytes than other phanerophytes. The other amphistomatous phanerophytes in this data set (\textit{Populus nigra} and \textit{Lavatera arborea}) are fast-growing pioneer species.

>>>>

% Things to do before submission:
% Remove code to copy refs
% word count
% plain language summary

@@

\usepackage{geometry}
\geometry{letterpaper}
\usepackage[parfill]{parskip}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{topcapt}
\usepackage{caption}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{natbib}
\usepackage{color}
\usepackage{array}
\usepackage{gensymb}
\usepackage{setspace}
\newcommand{\stretchy}{1.5}
\usepackage{lineno}
\usepackage{array}
\usepackage{lscape}
\usepackage{multirow}

@@

\title{Light and growth form interact to shape stomatal ratio among British angiosperms}
\author{Christopher D. Muir$^1$}
\date{} % delete this line to display the current date

@@

$^1$ Biodiversity Research Centre and Botany Department, University of British Columbia, Vancouver, British Columbia V6T 1Z4, Canada \\
\\
\textit{Author for correspondence:} \\
\textit{Christopher D. Muir} \\
\textit{Tel:} +17782284851 \\
\textit{Email:} chrisdmuir@gmail.com \\
University of British Columbia \\
6270 University Blvd. \\
Vancouver, BC, Canada \\
V6T 1Z4 \\
\\
Short title: Shedding light on stomatal evolution\\
\\
Word count: \\
Summary:  190\\
Introduction:  \\
Materials and Methods:  \\
Results:  \\
Discussion:  \\
Acknowledgement:  \\
3 Figures and 1 Table, 1 Supplemental Figure

@@

  # Remove before publication
  file.copy(from = "~/Google Drive/CommonResources/refs.bib",
            to = "~/Google Drive/britstom/ms/refs.bib", overwrite = TRUE)

@@

\begin{itemize}
	\item In most plants, stomata are located only on the abaxial leaf surface (hypostomy), but many plants have stomata on both surfaces (amphistomy). High light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together using phylogenetic comparative methods.
	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, Raunki\ae r lifeform, and phylogenetic relationships for \Sexpr{nrow(stomata)} species of British angiosperms. I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
	\item High light and herbaceous growth form are correlated with amphistomy, as predicted, but they also interact; the effect of light is pronounced in therophytes (annuals) and perennial herbs, but muted in phanerophytes (mostly trees). Interestingly, amphistomy and stomatal density evolve together in response to light, suggesting coordinated selection on this trait combination.
	\item I show for the first time that light and growth form interact to shape variation in stomatal ratio; amphistomy is advantageous in high light, but mostly for herbs. These results improve our understanding of the adaptive significance of stomatal ratio as well as its use as functional trait for paleoecology and crop improvement.
\end{itemize}

@@

Adaptation, amphistomy, Ellenberg light indicator value, growth form, phylogenetic comparative methods, Raunki\ae r lifeform, stomata, stomatal ratio

@@

The leading, nonmutually exclusive hypotheses are that 1) open habitats favour amphistomy because CO$_2$ diffusion most strongly limits photosynthetic rate under high light and 2) herbaceous growth form favours amphistomy because traits that maximize photosynthetic rate are often under stronger selection in herbs. \citeauthor{Salisbury_1927} (\citeyear{Salisbury_1927}) first noted that amphistomy is most common in herbaceous plants from open habitats (i.e., with high light) of the British flora. These observations have been replicated in other studies \citep{Mott_etal_1982, Peat_Fitter_1994b, Jordan_etal_2014, Muir_2015} and may support physiological and ecological hypotheses that CO$_2$ most strongly limits photosynthesis in high light and/or photosynthesis contributes most to fitness in herbaceous plants. Under high light, CO$_2$ can strongly limit maximum photosynthetic rates, espcecially in thick leaves \citep{Jones_1985}. Hence, having stomata on both surfaces relieves this limitation by adding a second parallel pathway for CO$_2$ diffusion. Parkhurst (\citeyear{Parkhurst_1978}) argued that greater leaf thickness \textit{per se} selected for amphistomy, but there is little evidence for correlations between leaf thickness and stomatal ratio independent of light \citep{Mott_etal_1982, Gibson_1996, Muir_2015}. Amphistomy is correlated with open habitat in warm desert plants of western North America \citep{Mott_etal_1982, Gibson_1996}, among the Proteaceae \citep{Jordan_etal_2014}, and in continental European herbs \citep{Bucher_etal_2017}.

Stomatal ratio is also associated with growth form. In the British flora, \citeauthor{Salisbury_1927} (\citeyear{Salisbury_1927}) found that trees and shurbs are nearly always hypostomatous, whereas herbs from open habitats are amphistomatous. This pattern holds when data are averaged by family to coarsely control for phylogenetic nonindependence \citep{Peat_Fitter_1994b} or when using alternative classification schemes, such as Raunki\ae r life form \citep{Peat_Fitter_1994b}. Across plants from $\sim 90$ familes worldwide, growth form is the strongest predictor of stomatal ratio when multiple factors are estimated simultaneously and controlling for phylogenetic nonindependence \citep{Muir_2015}. These patterns are consistent with other data indicating that many herbaceous plants are under strong selection for high maximum photosynthetic rates \citep{Bazzaz_1979, Korner_etal_1989, Wullschleger_1993}.

@@

The final question is important for addressing whether amphistomy is part of a coordinated syndrome of traits that promote higher photosynthetic rate, as both the light and growth form hypotheses assume. If evolved increases in stomatal ratio are mediated by shifting abaxial stomata to the adaxial surface, holding total stomatal density constant, then the overall increase in CO$_2$ diffusion would be small. In contrast, if amphistomy evolves by increasing adaxial stomatal density while holding abaxial density constant, then \textit{total} stomatal density must increase as well. Evolutionary coordination of amphistomy and high stomatal density would reinforce one another, increasing CO$_2$ supply to chloroplasts more than changes in either trait would in isolation.

@@

I obtained data on ab- and adaxial stomatal density on \Sexpr{nrow(stomata_unfiltered)} species from British Ecological Flora \citep{Salisbury_1927, Fitter_Peat_1994a, BEF}. Following recent comparative analyses \citep[e.g.][]{Bartelheimer_Poschlod_2016, Salguero-Gomez_etal_2016}, I used Ellenberg light indicator values \citep{Ellenberg_1974} and Raunki\ae r life form \citep{Raunkiaer_1934} as measures of light habitat and growth form, respectively. Hence, I am assuming that the species' light habitat is closely related to the type of habitat (open versus closed) where that species is found. Both attributes have been recently updated by taxonomic experts of the British flora (PLANTATT, \cite{Hill_etal_2004}). Ellenberg light indicator values are hereafter abbreviated \el. I used a dated molecular phylogeny of the British flora \citep{Lim_etal_2014} available from TreeBASE (http://treebase.org/; accession number 15105). \Sexpr{nNotInPhy1} species (\Sexpr{round(100 * nNotInPhy1 / nrow(stomata_unfiltered), 1)}\%) in the dataset were not present in the phylogeny. For \Sexpr{nNotInPhy1 - nNotInPhy2} of these species, I used the position a congeneric species as a proxy for the focal species \citep[following][]{Pennell_etal_2016}. When multiple congeneric species were present, I consulted the phylogenetic literature to identify the most closely related proxy species \citep{Scheen_etal_2004, Salmaki_etal_2013}. For the remaining \Sexpr{nNotInPhy2} missing species, I positioned them in the tree based on phylogenetic relationships to other genera or families present in the tree \citep{Fior_etal_2006}. Because many phylogenetic comparative methods do not allow polytomies, zero-length branches, and non-ultrametric trees, I made several small adjustments to the tree. I resolved polytomies randomly using the `multi2di' function in \pkg{phytools} version 0.6-00 \citep{Revell_2012}. I added 0.02 my to all zero-length branches, as this was approximately the length of the shortest nonzero branch length in the tree. After these changes, I slightly altered terminal branch lengths to make the tree precisely ultrametric.

@@

I compared phylogenetic linear models to test whether Raunki\ae r life form, \el, or interactions between them predicted $\textrm{SR}_\textrm{even}$. Unlike the analysis above, there was significant phylogenetic signal in this comparison (see R code). I used $\textrm{SR}_\textrm{even}$ rather than $\textrm{SR}_\textrm{propAd}$ as the response variable because the hypothesis is that faster life history and/or high light favor more even stomatal densities on each surface. I fit models using \pkg{phylolm} and extracted Akaike Information Criteria (AIC). For these and subsequent analyses, I assumed an Ornstein-Uhlenbeck process model for the residuals with the root character state integrated over the stationary distribution. I used a 10$^4$ parametric bootstrap samples of the full model (including main effects and interactions) to calculate parameter confidence intervals \citep{Boettiger_etal_2012}. % REMOVING, but keep text in case I decide to replace: Likewise, to determine whether the interaction between Raunki\ae r life form and \el~was statistically significant, I used a parametric bootstrap to generate the null distribution of $\Delta$AIC values ($\Delta$AIC is the difference in AIC between competing models). Specifically, I sampled 1000 random datasets from the estimated model with main effects of Raunki\ae r life form and \el~but no interaction. I fit these simulated datasets to models with and without interactions and  calculated $\Delta$AIC. The statistical significance of the observed $\Delta$AIC is the proportion of simulated $\Delta$AIC greater than the observed.

@@

I used two related phylogenetic methods, variance decomposition and structural equation modeling (SEM), to assess the relative contribution of ab- versus adaxial stomatal density to light-mediated stomatal ratio evolution. First, the contribution of ab- versus adaxial stomatal density can be calculated using phylogenetic variance decomposition methods as derived below. Because stomatal density is highly skewed, I log-transformed values for normality:

@@

I also wanted to test whether light-mediated evolution of stomatal ratio acted mostly by 1) increasing adaxial stomatal density while maintaining abaxial density, or 2) keeping total stomatal density the same, but shifting a greater proportion to the adaxial surface. The first scenario predicts that the phylogenetic regression of \el~ on $\mathrm{sd_{ad}}$ is stronger than that for $\mathrm{sd_{ab}}$. The second scenario predicts that \el~acts similarly on both and that there is a negative covariance ($\mathrm{Cov(sd_{ad}, sd_{ab}) < 0}$). I tested these competing predictions by fitting a very simple phylogenetic SEM \citep[see][for a similar approach]{Mason_etal_2016}. The model uses the phylogenetic covariance matrix, as described above, to simultaneously estimate regressions of \el~on $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ while allowing covariance between them (i.e. estimating $\mathrm{Cov(sd_{ad}, sd_{ab})}$). To fit the SEM, I used the R package \pkg{lavaan} version 0.5-23.1097 \citep{Rosseel_2012}. I tested whether parameter estimates were significantly different from zero using $z$-scores.

@@

The ratio of stomatal densities on the abaxial (`lower') to that of the adaxial (`upper') surface varies greatly across plant species, but the adaptive significance is not clear. Comparative studies correlating stomatal ratio to ecological factors can distinguish among competing hypotheses and reveal critical experiments for future work. Previous comparative studies suggested that high light and herbaceous growth form favor amphistomy \citep{Mott_etal_1982, Jordan_etal_2014, Muir_2015, Bucher_etal_2017}, particularly in the British flora \citep{Salisbury_1927, Peat_Fitter_1994b}. However, none of these studies have accounted for the fact that light and growth form are often confounded -- open, high light habitats are often dominated by herbs -- or the fact that species are not independent because of shared evolutionary hisory. Here, I reanalyzed data on stomata, light tolerance, and growth form in British angiosperms using phylogenetic comparative methods. As expected, species' light tolerance (Ellenberg light indicator or \el) is confounded with growth form (Raunki\ae r life form; Fig.~\ref{fig:lf-light}). Nevertheless, both \el~and Raunki\ae r life form affect stomatal ratio, but these factors also interact; the influence of \el~on stomatal ratio varies across forms. These novel findings provide further evidence that variation in stomatal ratio is adaptive and have important implications for interpreting changes in stomatal ratio through the paleo record \citep{Jordan_etal_2014} and during domestication \citep{Milla_etal_2013}.

@@

Previously, associations between light, growth form, and stomatal ratio have been interpreted in isolation as indicating that either high light and/or herbaceous growth form favors amphistomy. In British angiosperms, both factors are important, though statistical analyses suggest that light may be a stronger determinant than growth form (Table~\ref{table:srAIC}). Unlike previous studies, I found a significant interaction between light and growth form among British angiosperms, which suggests that amphistomy may only be strongly favored when CO$_2$ strongly limits photosynthesies (as in open habitat) \textit{and} photosynthesis strongly limits fitness (as in herbs). This is consistent with life history theory predicting that the demography of open habitat herbs is strongly limited by plant growth \citep{Franco_Silvertown_1996}. The ideal way to test this would be to measure selection on stomatal ratio in a species that varied quantitatively in both stomatal ratio and life history (e.g., containing both annual and perennial forms). I predict that amphistomy will be favored more strongly in the annual form grown under high light compared to an annual under low light or a perennial in high light, and much more strongly than a perennial grown in low light. Similar experiments could also be performed to test if and when light-mediated plasticity in stomatal ratio is adaptive \citep{Gay_Hurd_1975, Mott_Michaelson_1991, Fontana_etal_2017}.

@@

The weak or nonsignificant relationship between \el~and stomatal ratio in geophytes and phanerophytes suggests that in some cases amphistomy may not reliably indicate open habitat without further information. For example, perhaps amphistomatous geophytes from partially shaded habitats are spring ephemerals, so they  experience high light during their growth phase, but this has not been tested. Likewise, phanerophytes (most tall trees) are almost always hypostomatous (see also \cite{Muir_2015}). Most British phanerophytes are tall, hypostomatous trees, but the exceptions are telling. For example, the most amphistomatous phanerophyte in this dataset is \textit{Brassica oleracea}, a short-statured biennial that has more in common physiologically with hemicryptophytes than other phanerophytes. The other amphistomatous phanerophytes in this data set (\textit{Populus nigra} and \textit{Lavatera arborea}) are fast-growing pioneer species.

### 2017-07-14 ms/ms.tex
edited ms

\begin{itemize}
	\item In most plants, stomata are located only on the abaxial leaf surface (hypostomy), but many plants have stomata on both surfaces (amphistomy). High light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together using phylogenetic comparative methods.
	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, Raunki\ae r lifeform, and phylogenetic relationships for 372 species of British angiosperms. I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
	\item High light and herbaceous growth form are correlated with amphistomy, as predicted, but they also interact; the effects of light are most pronounced in therophytes (annuals) and perennial herbs, but less so in phanerophytes (mostly trees). Interestingly, amphistomy and stomatal density evolve together in response to light, suggesting coordinated selection on this trait combination.
	\item I show for the first time that light and growth form interact to shape variation in stomatal ratio; amphistomy is advantageous in high light, but mostly for herbs. These results improve our understanding of the adaptive significance of stomatal ratio as well as its use as functional trait for paleovegetation reconstruction and crop improvement.
\end{itemize}

@@

I obtained data on ab- and adaxial stomatal density on 395 species from British Ecological Flora \citep{Salisbury_1927, Fitter_Peat_1994a, BEF}. Following recent comparative analyses \citep[e.g.][]{Bartelheimer_Poschlod_2016, Salguero-Gomez_etal_2016}, I used Ellenberg light indicator values \citep{Ellenberg_1974} and Raunki\ae r life form \citep{Raunkiaer_1934} as measures of light habitat and growth form, respectively. Hence, I am assuming that the species' light habitat is closely related to the type of habitat (open versus closed) where that species is found. Both attributes have been recently updated by taxonomic experts of the British flora (PLANTATT, \cite{Hill_etal_2004}). Ellenberg light indicator values are hereafter abbreviated \el. I used a dated molecular phylogeny of the British flora \citep{Lim_etal_2014} available from TreeBASE (http://treebase.org/; accession number 15105). 14 species (3.5\%) in the dataset were not present in the phylogeny. For 8 of these species, I used the position a congeneric species as a proxy for the focal species. When multiple congeneric species were present, I consulted the phylogenetic literature to identify the most closely related proxy species \citep{Scheen_etal_2004, Salmaki_etal_2013}. For the remaining 6 missing species, I positioned them in the tree based on phylogenetic relationships to other genera or families present in the tree \citep{Fior_etal_2006}. Because many phylogenetic comparative methods do not allow polytomies, zero-length branches, and non-ultrametric trees, I made several small adjustments to the tree. I resolved polytomies randomly using the `multi2di' function in \pkg{phytools} version 0.6-00 \citep{Revell_2012}. I added 0.02 my to all zero-length branches, as this was approximately the length of the shortest nonzero branch length in the tree. After these changes, I slightly altered terminal branch lengths to make the tree precisely ultrametric.

>>>>

\begin{itemize}
	\item In most plants, stomata are located only on the abaxial leaf surface (hypostomy), but many plants have stomata on both surfaces (amphistomy). High light and herbaceous growth form have been hypothesized to favor amphistomy, but these hypotheses have not been rigourously tested together using phylogenetic comparative methods.
	\item I leveraged a large dataset including stomatal ratio, Ellenberg light indicator value, Raunki\ae r lifeform, and phylogenetic relationships for 372 species of British angiosperms. I used phylogenetic comparative methods to test how light and/or growth form influence stomatal ratio.
	\item High light and herbaceous growth form are correlated with amphistomy, as predicted, but they also interact; the effect of light is pronounced in therophytes (annuals) and perennial herbs, but muted in phanerophytes (mostly trees). Interestingly, amphistomy and stomatal density evolve together in response to light, suggesting coordinated selection on this trait combination.
	\item I show for the first time that light and growth form interact to shape variation in stomatal ratio; amphistomy is advantageous in high light, but mostly for herbs. These results improve our understanding of the adaptive significance of stomatal ratio as well as its use as functional trait for paleoecology and crop improvement.
\end{itemize}

@@

I obtained data on ab- and adaxial stomatal density on 395 species from British Ecological Flora \citep{Salisbury_1927, Fitter_Peat_1994a, BEF}. Following recent comparative analyses \citep[e.g.][]{Bartelheimer_Poschlod_2016, Salguero-Gomez_etal_2016}, I used Ellenberg light indicator values \citep{Ellenberg_1974} and Raunki\ae r life form \citep{Raunkiaer_1934} as measures of light habitat and growth form, respectively. Hence, I am assuming that the species' light habitat is closely related to the type of habitat (open versus closed) where that species is found. Both attributes have been recently updated by taxonomic experts of the British flora (PLANTATT, \cite{Hill_etal_2004}). Ellenberg light indicator values are hereafter abbreviated \el. I used a dated molecular phylogeny of the British flora \citep{Lim_etal_2014} available from TreeBASE (http://treebase.org/; accession number 15105). 14 species (3.5\%) in the dataset were not present in the phylogeny. For 8 of these species, I used the position a congeneric species as a proxy for the focal species \citep[following][]{Pennell_etal_2016}. When multiple congeneric species were present, I consulted the phylogenetic literature to identify the most closely related proxy species \citep{Scheen_etal_2004, Salmaki_etal_2013}. For the remaining 6 missing species, I positioned them in the tree based on phylogenetic relationships to other genera or families present in the tree \citep{Fior_etal_2006}. Because many phylogenetic comparative methods do not allow polytomies, zero-length branches, and non-ultrametric trees, I made several small adjustments to the tree. I resolved polytomies randomly using the `multi2di' function in \pkg{phytools} version 0.6-00 \citep{Revell_2012}. I added 0.02 my to all zero-length branches, as this was approximately the length of the shortest nonzero branch length in the tree. After these changes, I slightly altered terminal branch lengths to make the tree precisely ultrametric.
